Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-06-02 manuscript/manuscript.Rmd
minor updates.

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionalities simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better understand the range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.

Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018]. new toolkit such as BORIS were introduced [@friard2016]

@@

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is a power and elegant yet limited by the curse of expertise [@hinds1999]. Like chaine operatoire, it cannot handles variation very well. To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].

>>>>

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences invovled in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a power and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handles variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years. Here we use action grammar developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].
